---
title: 'Reproducible Research: Peer Graded Assignment'
author: "Ely Xavier"
date: "February 10, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the dataset

### 1) Loading the dataset from working directory
```{r loading}
if(!file.exists('activity.csv')){
        unzip('activity.zip')
}
data <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?

### 1) Calculate the total number of steps taken per day
```{r total_steps}
steps_by_day <- aggregate(steps ~ date, data, sum)
```

### 2) Make a histogram of the total number of steps taken each day 

```{r histogram}
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="gray", xlab="Number of Steps", breaks = 10)
```

### 3) Calculate and report the mean and median total number of steps taken per day

```{r mean_median}
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```
- Mean: r rmean
- Median: r rmedian

## What is the average daily activity pattern?
```{r pattern}
steps_by_interval <- aggregate(steps ~ interval, data, mean)
```

### 1) Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time_series}
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
```

### 2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r 5-min_interval}
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

- Most steps at: r max_intervals

## Imputing missing values

### 1) Calculate and report the total number of missing values in the dataset

```{r missing_values}
incomplete <- sum(!complete.cases(data))
```

- Number of missing values: r incomplete

### 2) Devise a strategy for filling in all of the missing values in the dataset.

### 3) Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{R imputed}
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

### 4) Make a histogram of the total number of steps taken each day

```{R histogram_imputed_values}
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="purple", xlab="Number of Steps", breaks = 10)

# Histogram to show differences
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="gray", xlab="Number of Steps", add=T, breaks = 10)
legend("topright", c("Imputed", "Non-imputed"), col=c("purple", "gray"), lwd=10)
```


### 5) Calcuulate and report the mean and median total number of steps taken in a day

```{R mean_and_median_imputed}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

- Mean (Imputed): r rmean.i
- Medain (Imputed): r rmedian.i

## Are there differences in activity patterns between weekdays and weekends?

### 1) Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r differences}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)
```

### 2) Make a panel plot containing a time series plot

```{r plot_differences}
library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```
